Graph Topological Aspects of Granger Causal Network Learning

### 2019-10-02 presentations/mcgillOct2019/pres.tex
Presentation almost done -- just fill in VAR(p) LASSO case.

\begin{frame}{Causality}
  \begin{itemize}
    \item{Do billiard collisions cause motion?}\pause
    \item{Does standing in the street cause cars to stop?}\pause
    \item{Does smoking cause cancer?}\pause
    \item{Does government spending cause an increase in GDP?}\pause
    \item{Granger causality attemps to address the last question's category.}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Basic Setup}
  \begin{itemize}
    \item{Basic setup is an $n-$dimensional WSS process $x(t)$}\pause
    \item{Wold representation $x(t) = \sum_{\tau = 0}^\infty A(\tau)v(t - \tau)$}\pause
    \item{$\text{VAR}(\infty)$ representation $x(t) = \sum_{\tau = 1}^\infty B(\tau)x(t - \tau) + v(t)$}\pause
    \item{Require $\E[v(t)v(t - \tau)] = \delta_{\tau}\Sigma_v$, with $\Sigma_v$ Diagonal}\pause
  \end{itemize}

@@

\begin{frame}{Granger Causality}
  \begin{defn}
    For the WSS series $x(t)$ we say that component $x_j$
    \textit{Granger-Causes} component $x_i$ (with respect to $x$)
    and write $x_j \gc x_i$ if given Hilbert spaces $\H_{t - 1}$,
    $\H^{-j}_{t - 1}$ if

@@

  \begin{defn}
    We say that $x_j$ pairwise Granger-causes $x_i$ and write
    $x_j \pwgc x_i$ if $x_j$ Granger-causes $x_i$ with respect only to
    $(x_i, x_j)$.
  \end{defn}
\end{frame}

@@

\begin{frame}{Simple Examples}
  \begin{align*}
    x_0(t) &= b_0 x_0(t - 1) + b_{0, 2} x_2(t - 150) + v_0(t)\\
    x_1(t) &= b_1 x_1(t - 1) + b_{1, 2} x_2(t - 150) + v_1(t)\\
    x_2(t) &= b_2 x_2(t - 1) + v_2(t)\\
  \end{align*}

  % TODO: Fix the whitespace consumed by this figure!

@@

\section{Causality Graphs}
\begin{frame}{Graph Terminology}
  \begin{itemize}
    \item{Graph $\gcg = (V, \gcge)$ with nodes (or vertices) $V$ and edges $\gcge$}\pause
    \item{Directed paths: $j \rightarrow a_0 \rightarrow \cdots \rightarrow a_r \rightarrow i$ for $(a_{k}, a_{k + 1}) \in \gcge$ distinct nodes}\pause
    \item{Cycles: paths with $i = j$}
    \item{Directed Acyclic Graph (DAG): A directed graph without any cycles}\pause
    \item{Parents: $j \in \pa{i}$ iff $(j, i) \in \gcge$}\pause
    \item{Ancestors: $j \in \anc{i}$ iff $j \rightarrow i$ path in $\gcge$}\pause
  \end{itemize}
\end{frame}

@@

\begin{frame}{Application Examples}
  Analysis of connectedness of financial institutions

  TODO: Finance example
  \begin{columns}

  \end{columns}

@@

  TODO: Anna examples or something
\end{frame}

\begin{frame}{Application Examples}
  TODO: Gene expression networks
\end{frame}

\section{Recovery Methods}
\begin{frame}{LASSO Recovery Methods}
  \begin{itemize}
    \item{Small systems ($n = 2$ or $n \lessapprox 10$) with adequate data are amenable to least squares and hypothesis testing}\pause
    \item{Large systems ($n \gtrapprox 20$) or when data is limited are addressed via the LASSO (and variants)}\pause
    \item{Standard linear regression case:
        \begin{equation*}
          \underset{\beta \in \R^n}{\text{minimize }} \frac{1}{2T}||y - X\beta||_2^2 + \lambda ||\beta||_1
        \end{equation*}
      }\pause
    \item{$\VAR(p)$ models:
        \begin{equation*}
          \underset{B(\tau) \in \R^{n \times n}}{\text{minimize }} \frac{1}{2T}||x(t) - \sum_{\tau = 1}^p B(\tau)x(t - \tau)||_2^2 + \lambda \sum_{\tau = 1}^p ||B(\tau)||_1
        \end{equation*}
      }
  \end{itemize}
\end{frame}

\begin{frame}{LASSO Recovery Theorems}
  TODO: Static Case
\end{frame}

\begin{frame}{LASSO Recovery Theorems}
  TODO: VAR(p) case
\end{frame}

\begin{frame}{Structure of Granger Causality Graphs}
  \begin{itemize}
    \item{Topology of Granger causality graph is a special feature of time series}\pause
    \item{TODO: Star shaped graph with a self-loop versus a full cycle with graph -- same number of edges, but qualitatively totally different.}
  \end{itemize}
\end{frame}
% \item{Confounder: A node $k$ confounds $i, j$ if $k \in \anc{i} \cap \anc{j}$ if exists a
%     path $\gcgpath{k}{i}$ not containing $j$, and a path
%     $\gcgpath{k}{j}$ not containing $i$.}

@@

\begin{frame}{Strongly Causal Graph Examples}
  \begin{columns}
    \begin{column}{0.33\linewidth}
      TODO: Simple example\pause
    \end{column}
    \begin{column}{0.33\linewidth}
      \small{Slightly larger example}
      \includegraphics[width=\linewidth]{../../figures/example_scg.pdf}\pause
    \end{column}
    \begin{column}{0.33\linewidth}
      \small{Similar number of edges, but not strongly causal}
      \includegraphics[width=\linewidth]{../../figures/example_dag.pdf}
    \end{column}
  \end{columns}
\end{frame}

@@

\begin{frame}{Pairwise Recovery}
  An Example (TODO: Intersperse Pauses)

  \begin{columns}
    \begin{column}{0.5\linewidth}
      \begin{figure}
        \includegraphics[width=\linewidth]{../../figures/example_algorithm.pdf}
      \end{figure}
    \end{column}

    \begin{column}{0.5\linewidth}
      \begin{figure}
        \includegraphics[width=\linewidth]{../../figures/example_algorithm2.pdf}
      \end{figure}
    \end{column}
  \end{columns}
\end{frame}

>>>>

\begin{frame}{Causality}
  \begin{itemize}
    \item{Do billiard collisions cause motion?}\pause
    \item{Does standing in the street cause cars to stop?}\pause
    \item{Does smoking cause cancer?}\pause
    \item{Does lowering interest rates cause lowering unemployment?}\pause
    \item{Granger causality attemps to address the last question's category.}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Basic Setup}
  \begin{itemize}
    \item{Basic setup is an $n-$dimensional WSS process $x(t)$}\pause
    \item{Wold representation $x(t) = \sum_{\tau = 0}^\infty A(\tau)v(t - \tau)$}\pause
    \item{$\VAR(\infty)$ representation $x(t) = \sum_{\tau = 1}^\infty B(\tau)x(t - \tau) + v(t)$}\pause
    \item{Require $\E[v(t)v(t - \tau)] = \delta_{\tau}\Sigma_v$, with $\Sigma_v$ Diagonal}\pause
  \end{itemize}

@@

\begin{frame}{Granger Causality}
  \begin{defn}
    For the WSS series $x(t)$ we say that component $x_j$
    \textit{Granger-Causes} component $x_i$ (with respect to $x$)
    and write $x_j \gc x_i$ if

@@

  \begin{defn}
    We say that $x_j$ pairwise Granger-causes $x_i$ and write
    $x_j \pwgc x_i$ if $x_j$ Granger-causes $x_i$ with respect to
    $(x_i, x_j)$.
  \end{defn}
\end{frame}

@@

\begin{frame}{Simple Examples}
  \begin{equation*}
    \begin{aligned}
      x_0(t) &= b_0 x_0(t - 1) + b_{0, 2} x_2(t - 150) + v_0(t)\\
      x_1(t) &= b_1 x_1(t - 1) + b_{1, 2} x_2(t - 150) + v_1(t)\\
      x_2(t) &= b_2 x_2(t - 1) + v_2(t)\\
    \end{aligned}
  \end{equation*}

@@

\section{Causality Graphs}
\begin{frame}{Graph Terminology}
  \begin{itemize}
    \item{Graph $\gcg = (V, \gcge)$ with nodes (or vertices) $V$ and edges $\gcge$}\pause
    \item{Directed paths: $j \rightarrow a_0 \rightarrow \cdots \rightarrow a_r \rightarrow i$ for $(a_{k}, a_{k + 1}) \in \gcge$ distinct nodes}\pause
    \item{Cycles: paths with $i = j$}
    \item{Directed Acyclic Graph (DAG): A directed graph without any cycles}\pause
    \item{Parents: $j \in \pa{i}$ iff $(j, i) \in \gcge$}\pause
    \item{Ancestors: $j \in \anc{i}$ iff $j \rightarrow i$ path in $\gcge$}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Granger Causality Graph}
  Structural results that tie in $\gcg$.  Recall:
  \begin{itemize}
    \item{$\mathsf{MA}(\infty)$ representation $x(t) = \sum_{\tau = 0}^\infty A(\tau)v(t - \tau)$}\pause
    \item{$\mathsf{VAR}(\infty)$ representation $x(t) = \sum_{\tau = 1}^\infty B(\tau)x(t - \tau) + v(t)$}\pause
  \end{itemize}

  \begin{proposition}
  $x_i(t)$ can be represented in terms of it's parents in $\gcg$:

  \begin{equation}
    \label{eqn:parent_expansion}
    x_i(t) = v_i(t) + \B_{ii}(z)x_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z)x_k(t).
  \end{equation}\pause

  Or, through it's ancestors:

  \begin{equation}
    \label{eqn:ancestor_expansion}
    x_i(t) = \A_{ii}(z)v_i(t) + \sum_{\substack{k \in \anc{i} \\ k \ne i}}\A_{ik}(z)v_k(t).
  \end{equation}

  \end{proposition}
\end{frame}

\begin{frame}{Application Examples}
  Analysis of connectedness of financial institutions: 2002 - 2004

  \begin{columns}
    \begin{column}{0.75\linewidth}
      \includegraphics[width=\linewidth]{./figures/finance_connections2002.png}
    \end{column}
  \end{columns}

\begin{frame}{Application Examples}
  Analysis of connectedness of financial institutions: 2006 - 2008

  \begin{columns}
    \begin{column}{0.75\linewidth}
      \includegraphics[width=\linewidth]{./figures/finance_connections2006.png}
    \end{column}
  \end{columns}

@@

  \begin{columns}
    \begin{column}{0.55\linewidth}
      \includegraphics[width=\linewidth]{./figures/brain_photo.png}
    \end{column}
  \end{columns}

  \footfullcite{anna_paper2008}
\end{frame}

\begin{frame}{Application Examples}
  Analysis of Gene regulatory networks

  \begin{columns}
    \begin{column}{0.5\linewidth}
      \includegraphics[width=\linewidth]{../../figures/ecoli_regulatory_network.pdf}
    \end{column}
  \end{columns}

  \footfullcite{discovering_graphical_Granger_causality_using_the_truncating_lasso_penalty}
\end{frame}

\section{Recovery Methods}
\begin{frame}{LASSO Recovery Methods}
  \begin{itemize}
    \item{Small systems ($n = 2$ or $n \lessapprox 10$) with adequate data are amenable to least squares and hypothesis testing}\pause
    \item{Large systems ($n \gtrapprox 20$) or when data is limited are addressed via the LASSO (and variants)}\pause
    \item{Standard linear regression case $y = X\beta^\star + v$:
        \begin{equation*}
          \underset{\beta \in \R^n}{\text{minimize }} \frac{1}{2T}||y - X\beta||_2^2 + \lambda ||\beta||_1
        \end{equation*}
      }\pause
    \item{$\VAR(p)$ models $x(t) = \sum_{\tau = 1}^p B^\star(\tau)x(t - \tau) + v(t)$:
        \begin{equation*}
          \underset{B(\tau) \in \R^{n \times n}}{\text{minimize }} \frac{1}{2T}||x(t) - \sum_{\tau = 1}^p B(\tau)x(t - \tau)||_2^2 + \lambda \sum_{\tau = 1}^p ||B(\tau)||_1
        \end{equation*}
      }
  \end{itemize}
\end{frame}

\begin{frame}{LASSO Recovery Theorems (Linear Regression)}
  \begin{itemize}
    \item{$y = X\beta^\star + v$; $\beta^\star \in \R^n, y \in \R^T$}\pause
    \item{$\widehat{\beta} \in \underset{\beta \in \R^n}{\text{argmin }} \frac{1}{2T}||y - X\beta|| + \lambda ||\widehat{\beta}||_1$}\pause
  \end{itemize}

  \begin{theorem}[Wainwright 2006]
    Suppose $\beta^\star$ has only $s$ non-zero terms:
    $||\beta^\star||_0 = s$, and $v \sim \mathcal{N}(0, \sigma^2)$.
    \pause Then, with high probability, (ignoring some important
    technicalities) there is a constant $C$ s.t. \pause

    \begin{equation*}
      ||\widehat{\beta} - \beta^\star||_2 \le C \sigma \sqrt{\frac{s \mathsf{log } n}{T}}.
    \end{equation*}
  \end{theorem}

  \begin{itemize}
    \item{Roughly, $||\beta^\star - \widehat{\beta}||_2$ is $O\big(\frac{s}{T}\big)$.}
  \end{itemize}
\end{frame}

\begin{frame}{LASSO Recovery Theorems ($\VAR(p)$ case)}
  \begin{itemize}
    \item{$\frac{1}{2T}||x(t) - \sum_{\tau = 1}^p B(\tau)x(t - \tau)||_2^2 + \lambda \sum_{\tau = 1}^p ||B(\tau)||_1$}\pause
  \end{itemize}
  TODO
\end{frame}

\begin{frame}{Structure of Granger Causality Graphs}
  \begin{columns}
    \begin{column}{0.35\linewidth}
      \includegraphics[width=\linewidth]{../../figures/example_rotation_shape.pdf}\pause
    \end{column}
    \begin{column}{0.35\linewidth}
      \includegraphics[width=\linewidth]{../../figures/example_star_shaped.pdf}\pause
    \end{column}
  \end{columns}

  \begin{columns}
    \begin{column}{0.35\linewidth}
      \includegraphics[width=\linewidth]{../../figures/example_loop_shaped.pdf}\pause
    \end{column}
    \begin{column}{0.35\linewidth}
      \includegraphics[width=\linewidth]{../../figures/example_diamond_shaped.pdf}
    \end{column}
  \end{columns}
\end{frame}
% \item{Confounder: A node $k$ confounds $i, j$ if $k \in \anc{i} \cap \anc{j}$ if exists a
%     path $\gcgpath{k}{i}$ not containing $j$, and a path
%     $\gcgpath{k}{j}$ not containing $i$.}

@@

\begin{frame}{Strongly Causal Graph Examples}
  \begin{columns}
    \begin{column}{0.33\linewidth}
      \small{Simple Example}
      \includegraphics[width=\linewidth]{../../figures/example_algorithm_step0.pdf}\pause
    \end{column}
    \begin{column}{0.33\linewidth}
      \small{Slightly larger example}
      \includegraphics[width=\linewidth]{../../figures/example_scg.pdf}\pause
    \end{column}
    \begin{column}{0.33\linewidth}
      \small{Similar number of edges, but not strongly causal}
      \includegraphics[width=\linewidth]{../../figures/example_dag.pdf}
    \end{column}
  \end{columns}
\end{frame}

@@

\begin{frame}{Pairwise Recovery}
  \begin{columns}
    \begin{column}{0.5\linewidth}
      \begin{figure}
        \includegraphics[width=\linewidth]{../../figures/example_algorithm.pdf}
      \end{figure}
    \end{column}

    \begin{column}{0.5\linewidth}
      \begin{figure}
        \includegraphics[width=\linewidth]{../../figures/example_algorithm_step0.pdf}
      \end{figure}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Pairwise Recovery}
  \begin{columns}
    \begin{column}{0.5\linewidth}
      \begin{figure}
        \includegraphics[width=\linewidth]{../../figures/example_algorithm0_step1.pdf}
      \end{figure}
    \end{column}

    \begin{column}{0.5\linewidth}
      \begin{figure}
        \includegraphics[width=\linewidth]{../../figures/example_algorithm_step0.pdf}
      \end{figure}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Pairwise Recovery}
  \begin{columns}
    \begin{column}{0.5\linewidth}
      \begin{figure}
        \includegraphics[width=\linewidth]{../../figures/example_algorithm0_step1.pdf}
      \end{figure}
    \end{column}

    \begin{column}{0.5\linewidth}
      \begin{figure}
        \includegraphics[width=\linewidth]{../../figures/example_algorithm_step1.pdf}
      \end{figure}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Pairwise Recovery}
  \begin{columns}
    \begin{column}{0.5\linewidth}
      \begin{figure}
        \includegraphics[width=\linewidth]{../../figures/example_algorithm0_step1.pdf}
      \end{figure}
    \end{column}

    \begin{column}{0.5\linewidth}
      \begin{figure}
        \includegraphics[width=\linewidth]{../../figures/example_algorithm_step2.pdf}
      \end{figure}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Pairwise Recovery}
  \begin{columns}
    \begin{column}{0.5\linewidth}
      \begin{figure}
        \includegraphics[width=\linewidth]{../../figures/example_algorithm0_step1.pdf}
      \end{figure}
    \end{column}

    \begin{column}{0.5\linewidth}
      \begin{figure}
        \includegraphics[width=\linewidth]{../../figures/example_algorithm_step3.pdf}
      \end{figure}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Pairwise Recovery}
  \begin{columns}
    \begin{column}{0.5\linewidth}
      \begin{figure}
        \includegraphics[width=\linewidth]{../../figures/example_algorithm0_step1.pdf}
      \end{figure}
    \end{column}

    \begin{column}{0.5\linewidth}
      \begin{figure}
        \includegraphics[width=\linewidth]{../../figures/example_algorithm_step4.pdf}
      \end{figure}
    \end{column}
  \end{columns}
\end{frame}
